---
title: "Genetic Structure"
output: rmdformats::robobook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, message = FALSE)

library(dplyr)
library(ggplot2)
library(vroom)
library(stringr)
library(here)
library(ape)
library(ggpubr)
library(knitr)

sample_info <- vroom(here("sample_info/internal_external_sample_info.tsv"))
```

# Data set structure

Samples were collected by the SOSFERT project, plus external data to increase the sample size and the number of breeds. A total of 104 samples were used for SNP calling.

```{r}
sample_info %>% 
  group_by(breed, project) %>% 
  summarise(n = n()) %>% 
  reshape2::dcast(project ~ breed, value.var = "n") %>% 
  janitor::adorn_totals(where = c("row","col")) %>% 
  kable()
```


# Principal Component Analysis 

```{r}

pc_pct <- here("11_genetic_str/output/pc_pct.csv") %>% read.csv()

pc_eigenscores <- here("11_genetic_str/output/pc_eigenscores.csv") %>% read.csv() %>% 
  inner_join(sample_info, by = c("sample.id"="sample_id"))

```

### Scree Plot 

The first PC contains ~9% of the variance. PC1-4 together amount to ~25% of the variance. The first 7 PCs have ~30% of the variance.

```{r, fig.height=5, fig.width=5} 
pc_pct %>% 
  ggplot(aes(x=PC,y=varprop)) +
    geom_bar(stat = "identity") +
    xlab("Principal Component") +
    ylab("% Variance") +
    scale_x_continuous(breaks = 1:7) +
    theme_bw()
```


### PCA plots colored by breed

Breeds are well clustered along PC1-2, PC3-4. However, along PC5 samples breeds cannot be differentiated anymore, and along PC6 European wild boars depart from the other breeds.
 
```{r, fig.height=3, fig.width=8}
p1 <- pc_eigenscores %>% 
        ggplot(data = ., aes(x=PC1,y=PC2,color=breed)) +
          geom_point() +
          theme_bw(base_size = 12) +
          theme(legend.title = element_blank())

p2 <- pc_eigenscores %>% 
        ggplot(data = ., aes(x=PC3,y=PC4,color=breed)) +
          geom_point() +
          theme_bw(base_size = 12) +
          theme(legend.title = element_blank())

p3 <- pc_eigenscores %>% 
        ggplot(data = ., aes(x=PC5,y=PC6,color=breed)) +
          geom_point() +
          theme_bw(base_size = 12) +
          theme(legend.title = element_blank())



#ggarrange(p1,p4, p2,p5, p3,p6, ncol = 2, nrow = 3)


ggarrange(p1,p2, p3, ncol = 3, nrow = 1, common.legend = TRUE)



```

### PCA plots colored by project

Projects seem to be well mixed, consider that each breed is represented by only a few projects. 

```{r, fig.height=3, fig.width=8}

p4 <- pc_eigenscores %>% 
        ggplot(data = ., aes(x=PC1,y=PC2,color=project)) +
          geom_point() +
          theme_bw(base_size = 12) +
          theme(legend.title = element_blank())

p5 <- pc_eigenscores %>% 
        ggplot(data = ., aes(x=PC3,y=PC4,color=project)) +
          geom_point() +
          theme_bw(base_size = 12) +
          theme(legend.title = element_blank())

p6 <- pc_eigenscores %>% 
        ggplot(data = ., aes(x=PC5,y=PC6,color=project)) +
          geom_point() +
          theme_bw(base_size = 12) +
          theme(legend.title = element_blank())

ggarrange(p4,p5, p6, ncol = 3, nrow = 1, common.legend = TRUE)

```


# Hierarchical Clustering

There are five clusters, one for every breed.

```{r prepare_dendrogram_data, include=FALSE}
dend <- readRDS(here("11_genetic_str/output/dendrogram_full_snp_set.rds")) %>% as.hclust()

# change labels to group to avoid legend, re-order labels
sample_info_hc_sorted <- sample_info[match(dend[["labels"]], sample_info$sample_id),]
identical(sample_info_hc_sorted$sample_id, dend[["labels"]]) #check order of labels is the same! yes!

dend[["labels"]] <- sample_info_hc_sorted$breed
```


```{r} 
# colors based on palette Dark2 used later for admixture
colrs <- c("#7570B3","#66A61E", "#1B9E77", "#D95F02", "#E7298A")
names(colrs) <- sample_info_hc_sorted$breed %>% unique()

# make 5 clusters
clus5 <- cutree(dend, 5)

# add colors by cluster, matching line with corresponding color
colrs_tip <- colrs[names(clus5)]

# Shorten label
dend[["labels"]][dend[["labels"]] == "European Wild Boar"] <- "EWB"
dend[["labels"]][dend[["labels"]] == "Pietrain"] <- "PI"
dend[["labels"]][dend[["labels"]] == "Large White"] <- "LW"
dend[["labels"]][dend[["labels"]] == "Landrace"] <- "LR"
dend[["labels"]][dend[["labels"]] == "Duroc"] <- "DU"

plot(as.phylo(dend), 
     type = "fan", 
     cex = 0.6,
     no.margin = TRUE,
     tip.color = colrs_tip)

# good reference: http://www.sthda.com/english/wiki/beautiful-dendrogram-visualizations-in-r-5-must-known-methods-unsupervised-machine-learning
```

# ADMIXTURE

Starting from two ancestral population, EWB has an equal ancestry proportion. Breeds are represented almost uniquely by one ancestor, with very little admixture.

As the number of ancestral populations increase, so does the level of admixture in landrance. However, the rest of breeds and EWB display very little admixture.

EWB did not represent breeds as expected. 


```{r import_fam_info}
# About sample ordering: https://www.biostars.org/p/221817/

fam_fl <- read.table(here("11_genetic_str/output/cohort_biallelicSNPs_HardFiltered_WithMissingness_filtered_header_fixed.fam"))
```

```{r make_ancestry_barplot_function}
make_ancestry_barplot <- function(k){

#k=2
  
  q_fl=here(paste0("11_genetic_str/output/cohort_biallelicSNPs_HardFiltered_WithMissingness_filtered_header_fixed.",k,".Q"))
  
  k_dat <- read.table(q_fl) %>% 
    bind_cols(
      dplyr::select(fam_fl, V2) %>% dplyr::rename(sample_id = V2)
      ) %>% 
    left_join(sample_info, by = "sample_id") %>% 
    reshape2::melt(id.vars = c("sample_id","breed","project")) %>% 
    mutate(breed = factor(breed, levels = c("European Wild Boar","Large White","Landrace","Pietrain","Duroc"))) %>% 
    arrange(breed, sample_id)
  
  k_dat %>% 
    ggplot(aes(x = sample_id, y = value,fill = variable)) +
    geom_bar(stat = "identity", position = "stack", width = 1) +
    theme_bw(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      legend.position = "none",
      plot.title = element_text(hjust = 0.5)
    ) +
    xlab(NULL) +
    ylab("Ancestry Fraction") +
    ggtitle(paste0("K=",k)) +
    scale_y_continuous(expand = c(0.005,0.005), breaks = seq(0,1,0.2)) +
    scale_fill_brewer(palette = "Dark2") +
    facet_grid(~breed, scales = "free") +
    theme(panel.spacing = unit(1/15, "lines"),
          strip.background = element_blank(), 
          panel.border = element_blank(),
          panel.grid = element_blank(),
          axis.text.x = element_blank()) 
  }

```


### Ancestry fraction (k2 - k6)
```{r, fig.height=10, fig.width=7}
ggarrange(make_ancestry_barplot(2),
          make_ancestry_barplot(3),
          make_ancestry_barplot(4),
          make_ancestry_barplot(5),
          make_ancestry_barplot(6), ncol = 1, nrow = 5)
```






