---
title: "Shared Genetic Differentiation from EWB"
output: rmdformats::robobook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, message = FALSE)

library(here)
library(vroom)
library(dplyr)
library(knitr)
library(ggpubr)
library(GenomicRanges)
library(WebGestaltR)
```

# Intro
Regions of extreme genetic differentiation were previously detected between dam or sire gorups relative to european wild boar.

These analysis were done in separate reports, for clarity.

This document compares the outlier regions detected for dam and sire.

# Shared genes in outlier Regions

```{r import_gene_lists}

genes_outliers_fst_dam_ewb <- here("00_analysis/data/genes_outliers_fst_dam_ewb.csv") %>% vroom()

genes_outliers_fst_sire_ewb <- here("00_analysis/data/genes_outliers_fst_sire_ewb.csv") %>% vroom()

```

```{r find_shared_genes}

shared_genes <- inner_join(genes_outliers_fst_dam_ewb %>% dplyr::select(seqnames,start, end, gene_id, gene_name, gene_biotype), 
                           genes_outliers_fst_sire_ewb %>% dplyr::select(seqnames,start, end, gene_id, gene_name, gene_biotype),
                           by = c("seqnames","start", "end", "gene_id","gene_name", "gene_biotype")) %>% 
  arrange(seqnames, start, end)

```

There are `r nrow(shared_genes)` shared genes in outliers regions detected for dam and sire groups.

```{r display_table}

shared_genes %>% DT::datatable(rownames = FALSE, options = list(pageLength = 40))

```

# Enrichment analysis

No significant gene set is identified based on FDR 0.1!

```{r function_run_WebGestaltR, eval = FALSE} 
#---------------------------
# function_run_WebGestaltR
#---------------------------

run_WebGestaltR <- function(
  interestGene, # vector of genes as input
  projectName, # the suffix to which "Project_" is appended
  enrichDatabase, # the data base to query  
  outputDirectory # where to store the results
  ){
  res <- WebGestaltR(
    enrichDatabase = enrichDatabase,
    enrichMethod = "ORA",
    organism = "sscrofa",
    interestGene = interestGene,
    interestGeneType = "ensembl_gene_id",
    referenceSet = "genome",
    sigMethod = "fdr",
    fdrMethod = "BH",
    fdrThr = 0.1,
    topThr = 100,
    reportNum = 20,
    isOutput = TRUE,
    outputDirectory = outputDirectory,
    hostName = "http://www.webgestalt.org/",
    projectName = projectName
  )
  
  write.csv(
    res, 
    file.path(outputDirectory,paste0("Project_", projectName), paste0(enrichDatabase,"_sig_results.csv"))
  )
}

```

```{r run_WebGestaltR, eval = FALSE} 
run_WebGestaltR(
  interestGene=unique(shared_genes$gene_id),
  projectName="GOBP",
  enrichDatabase="geneontology_Biological_Process",
  outputDirectory=here("00_analysis/data/WebGestaltR_shared_fst_vs_ewb")
)


run_WebGestaltR(
  interestGene=unique(shared_genes$gene_id),
  projectName="KEGG",
  enrichDatabase="pathway_KEGG",
  outputDirectory=here("00_analysis/data/WebGestaltR_shared_fst_vs_ewb")
)

```