---
title: "Gene Inspection"
output: 
  rmdformats::robobook:
    toc_depth: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, message = FALSE)
library(dplyr)
library(here)
library(vroom)
library(knitr)
library(stringr)
```

```{r import_gene_lists} 

genes_outliers_fst_dam_sire <- read.csv(here("00_analysis/data/genes_outliers_fst.csv")) %>% as_tibble()

genes_outliers_sire_pi_ratio <- read.csv(here("00_analysis/data/genes_outliers_sire_pi_ratio.csv")) %>% as_tibble()

genes_outliers_dam_pi_ratio <- read.csv(here("00_analysis/data/genes_outliers_dam_pi_ratio.csv")) %>% as_tibble()

genes_outliers_fst_dam_ewb <- here("00_analysis/data/genes_outliers_fst_dam_ewb.csv") %>% vroom()

genes_outliers_fst_sire_ewb <- here("00_analysis/data/genes_outliers_fst_sire_ewb.csv") %>% vroom()


# combine gene lists
gene_lists <- list(fst_dam_sire = genes_outliers_fst_dam_sire,
     sire_pi_ratio = genes_outliers_sire_pi_ratio,
     dam_pi_ratio = genes_outliers_dam_pi_ratio, 
     fst_dam_ewb = genes_outliers_fst_dam_ewb,
     fst_sire_ewb = genes_outliers_fst_sire_ewb) %>% 
  bind_rows(.id = "list") %>% 
  dplyr::select(list, gene_id, gene_name) %>% 
  mutate(list = factor(list, levels = c("fst_dam_sire","fst_dam_ewb","fst_sire_ewb","sire_pi_ratio","dam_pi_ratio")))

gns <- gene_lists$gene_id %>% unique()

```

```{r add_GOBP_and_export, eval = FALSE}
library(biomaRt)

mart <- useDataset(dataset = "sscrofa_gene_ensembl", mart = useMart("ENSEMBL_MART_ENSEMBL", host = "www.ensembl.org"))

gobp_terms <- getBM(attributes = c("ensembl_gene_id","external_gene_name","name_1006","namespace_1003","go_id"),
                              filters = "ensembl_gene_id", 
                              values = gns,
                              mart = mart) %>% 
  filter(namespace_1003 == "biological_process") %>% 
  mutate(gobp_goid = paste0(name_1006,"(",go_id,")")) #1539    6

write.csv(gobp_terms, here("00_analysis/data/genes_of_interest_and_gobp.csv"))

```

```{r add_KEGG_and_export, eval = FALSE}
library(KEGGREST) #BiocManager::install("KEGGREST")

# convert ensembl ids to entrez ids
ensembl_to_entrez <- getBM(attributes = c("ensembl_gene_id", "entrezgene_id"),#"kegg_enzyme"),
      filters = "ensembl_gene_id", 
      values = gns,
      mart = mart) 

ensembl_to_entrez <- ensembl_to_entrez %>% 
  filter(!is.na(entrezgene_id)) %>% 
  mutate(ssc_entrezgene_id = paste0("ssc:",entrezgene_id))

# genes to query
tmp <- unique(ensembl_to_entrez$ssc_entrezgene_id) #values: gene ensembl ids

# indixes to split data
# only 10 genes at a time are allowed
length(tmp)
query_df <- data.frame(start = seq(1, length(tmp), 10)) %>% mutate(end = start + 9)
head(query_df)
tail(query_df)

# query in batches of max 10 genes
KEGGs <- lapply(1:nrow(query_df), function(i){
  
  print(i)

  query_genes <- tmp[query_df$start[i]:query_df$end[i]]

  kegg_res <- keggGet(query_genes) %>% 
    lapply(function(x){ 
      data.frame(pathway_id = names(x$PATHWAY) , pathway = x$PATHWAY ) %>% 
        mutate(entrez_gene = x$ENTRY %>% as.integer()) %>% 
        dplyr::select(entrez_gene, everything())
      }) %>% 
    bind_rows()
  
  if(nrow(kegg_res) == 0){
    print("Neither gene matched KEGG terms")
  }else{
    kegg_res %>% 
      inner_join(ensembl_to_entrez, by = c("entrez_gene"="entrezgene_id")) 
  }
  }) 

# export results
KEGGs %>% bind_rows() %>% unique() %>% as_tibble() %>% dplyr::select(-mmu_entrezgene_id) %>% write.csv("00_analysis/data/genes_of_interest_and_KEGG.csv")
```

```{r gene_gene_description, eval = FALSE}

gn_descr <-  getBM(attributes = c("ensembl_gene_id","external_gene_name","description"),
                  filters = "ensembl_gene_id", 
                  values = gns,
                  mart = mart)

write.csv(gn_descr, here("00_analysis/data/genes_of_interest_and_description.csv"))

```


```{r import_GOBP_and_KEGG_annotations, include=FALSE}

GOBPs <- here("00_analysis/data/genes_of_interest_and_gobp.csv") %>% read.csv() %>% as_tibble() %>% dplyr::select(-X)

KEGGs <- here("00_analysis/data/genes_of_interest_and_KEGG.csv") %>% read.csv() %>% as_tibble() %>% dplyr::select(-X)

gene_description <- here("00_analysis/data/genes_of_interest_and_description.csv") %>% read.csv() %>% as_tibble() %>% dplyr::select(-X)
```



# Intro

This documents collects information for each gene detected in outlier regions for:

* Fst Dam vs Sire

* Pi ratio low diversity in dam

* Pi ratio low diversity in sire

* Fst Dam vs wild boar

* Fst Sire vs wild boar

Information for each gene was collected mostly from ensembl, ncbi-gene and pubmed.

```{r}
gene_lists %>% 
  group_by(list) %>% 
  summarise(n = n()) %>% 
  kable(caption = "Number of genes per list")
```

# Usefull information and lists

## Genes found in more than one list

Genes detected by different methods to evaluate divergence between genomes of dam and sire gorups.

```{r, include=FALSE}
# find genes found in more than one list
shared_genes <- gene_lists %>% 
  group_by(gene_id, gene_name) %>% 
  summarise(n=n()) %>% 
  filter(n > 1)
  
# display repeated genes
shared_genes_wide <- gene_lists %>% 
  reshape2::dcast(gene_id + gene_name ~ list, fun.aggregate = length) %>% 
  filter(gene_id %in% shared_genes$gene_id) %>% 
  as_tibble() 

shared_genes_wide[,3:7] %>% apply(1,sum) %>% max() # genes appear repeated in max 2 lists.


```

There are `r gns %>% length()` genes in total, Of which `r nrow(shared_genes)` were found by more than one method. 

Genes appear repeated in no more than 2 lists.

For each pair of lists genes were analyzed.

```{r display_shared_genes}
shared_genes_wide %>% DT::datatable(rownames = FALSE, options = list(pageLength = 20), caption = "Gene ids and corresponding lists")
```

```{r, fig.width=5, fig.height=8}
pheatmap::pheatmap(shared_genes_wide[,3:7], show_rownames = FALSE, 
                   main = "Visual representation of gene overlaps among lists", 
                   color = c("lightgrey","black"), legend_breaks = c(0,1))
```

## Shared GOBPs

Gene Ontologies Biological Process shared by at least 2 genes.

```{r}

shared_gobps <- GOBPs %>% 
  mutate(gene_name = ifelse(external_gene_name == "", ensembl_gene_id, external_gene_name)) %>% 
  group_by(go_id, name_1006) %>% 
  summarise(n = n(), genes = paste(gene_name, collapse = " / ")) %>% 
  filter(n > 1) %>% 
  arrange(desc(n))

shared_gobps %>% DT::datatable(rownames = FALSE, options = list(pageLenth = 20))


```

## Shared KEGGs

KEGGs shared by at least 2 genes.

```{r}

shared_keggs <- KEGGs %>% 
  # add gene names
  full_join(gene_lists, by = c("ensembl_gene_id"="gene_id")) %>% 
  mutate(gene_name = ifelse(is.na(gene_name), ensembl_gene_id, gene_name)) %>% 
  filter(!is.na(pathway)) %>% 
  dplyr::group_by(pathway) %>% 
  summarise(n = n(), genes = paste(gene_name, collapse = " / ")) %>% 
  filter(n > 1) %>% 
  arrange(desc(n))

shared_keggs %>% DT::datatable(rownames = FALSE, options = list(pageLenth = 20))

```



