---
title: "Genetic Structure - SOSFERT samples"
output: 
  html_document:
    toc: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(vroom)
library(ape)
library(ggpubr)
```

```{r}

sample_info <- vroom(here("sample_info/internal_external_sample_info.tsv"))

```

# Intro

Genetic structure of animals sequenced at high depth (~50x) as part of the original conception of the SOSFERT project were analysed to characterize the genetic structure in the combined population of Large White and Pietrain.

For this, the whole SNP was used (~16M) to compute principal component analysis, hierarchical clustering and admixture. The results can be seen below. 


# Principal Component Analysis

Detection of clusters of animals using.

```{r}

pc_pct <- here("11_genetic_str/output/pc_pct_SOSFERTsamples.csv") %>% read.csv()

pc_eigenscores <- here("11_genetic_str/output/pc_eigenscores_SOSFERTsamples.csv") %>% 
  read.csv() %>% 
  inner_join(sample_info, by = c("sample.id"="sample_id"))

```

### Scree Plot 

The scree plot shows that most of the variance is found in the first principal component.

```{r, fig.height=5, fig.width=5} 
pc_pct %>% 
  ggplot(aes(x=PC,y=varprop)) +
    geom_bar(stat = "identity") +
    xlab("Principal Component") +
    ylab("% Variance") +
    scale_x_continuous(breaks = 1:7) +
    theme_bw()
```

### PCA plots colored by breed

The scatter plot of eigen scores shows a clear separation between pietrain and large-white samples.

```{r, fig.height=5, fig.width=8}
pc_eigenscores %>% 
  ggplot(data = ., aes(x=PC1,y=PC2,color=breed)) +
    geom_point() +
    theme_bw(base_size = 12) +
    theme(legend.title = element_blank())
```

# Hierarchical Clustering

Hierarchical clustering was computed based on the genetic similarity of samples (identity by state). As the PCA, HC also distinguishes two clear populations.

```{r prepare_dendrogram_data, include=FALSE}
dend <- readRDS(here("11_genetic_str/output/dendrogram_full_snp_set_SOSFERTsamples.rds")) %>% as.hclust()

# change labels to group to avoid legend, re-order labels
sample_info_hc_sorted <- sample_info[match(dend[["labels"]], sample_info$sample_id),]
identical(sample_info_hc_sorted$sample_id, dend[["labels"]]) #check order of labels is the same! yes!

dend[["labels"]] <- sample_info_hc_sorted$breed
```

```{r} 
# colors based on palette Dark2 used later for admixture
#colrs <- c("#7570B3","#66A61E", "#1B9E77", "#D95F02", "#E7298A")
colrs <- c("#7570B3","#E7298A")
names(colrs) <- sample_info_hc_sorted$breed %>% unique()

# make 5 clusters
clus5 <- cutree(dend, 5)

# add colors by cluster, matching line with corresponding color
colrs_tip <- colrs[names(clus5)]

# Shorten label
dend[["labels"]][dend[["labels"]] == "Pietrain"] <- "PI"
dend[["labels"]][dend[["labels"]] == "Large White"] <- "LW"

plot(as.phylo(dend), 
     type = "fan", 
     cex = 0.6,
     no.margin = FALSE,
     tip.color = colrs_tip)

# good reference: http://www.sthda.com/english/wiki/beautiful-dendrogram-visualizations-in-r-5-must-known-methods-unsupervised-machine-learning

```


# ADMIXTURE

Admixture analysis shows that if two past populations are considered, there is no evidence of admixture in the present. 

More than three populations shows (K=3 - K=6) substructures in each breed, but no admixture.

```{r import_fam_info}
# About sample ordering: https://www.biostars.org/p/221817/

fam_fl <- read.table(here("11_genetic_str/output/cohort_biallelicSNPs_HardFiltered_WithMissingness_SOSFERTsamples_GTfiltered.fam"))
```

```{r make_ancestry_barplot_function}
make_ancestry_barplot <- function(k){

#k=2
  
  q_fl=here(paste0("11_genetic_str/output/cohort_biallelicSNPs_HardFiltered_WithMissingness_SOSFERTsamples_GTfiltered.",k,".Q"))
  
  k_dat <- read.table(q_fl) %>% 
    bind_cols(
      dplyr::select(fam_fl, V2) %>% dplyr::rename(sample_id = V2)
      ) %>% 
    left_join(sample_info, by = "sample_id") %>% 
    reshape2::melt(id.vars = c("sample_id","breed","project")) %>% 
    mutate(breed = factor(breed, levels = c("Large White", "Pietrain"))) %>% 
    arrange(breed, sample_id)
  
  k_dat %>% 
    ggplot(aes(x = sample_id, y = value,fill = variable)) +
    geom_bar(stat = "identity", position = "stack", width = 1) +
    theme_bw(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      legend.position = "none",
      plot.title = element_text(hjust = 0.5)
    ) +
    xlab(NULL) +
    ylab("Ancestry Fraction") +
    ggtitle(paste0("K=",k)) +
    scale_y_continuous(expand = c(0.005,0.005), breaks = seq(0,1,0.2)) +
    scale_fill_brewer(palette = "Dark2") +
    facet_grid(~breed, scales = "free") +
    theme(panel.spacing = unit(1/15, "lines"),
          strip.background = element_blank(), 
          panel.border = element_blank(),
          panel.grid = element_blank(),
          axis.text.x = element_blank()) 
  }

```

### Ancestry fraction (k2 - k6)
```{r, fig.height=10, fig.width=7}
ggarrange(make_ancestry_barplot(2),
          make_ancestry_barplot(3),
          make_ancestry_barplot(4), 
          make_ancestry_barplot(5), 
          make_ancestry_barplot(6),
          ncol = 1, nrow = 5)
```
