---
title: "Intersection between (RED genes DU mouse) x (Fst Outliers genes SOSFERT pig)"
output: 
  html_document:
    toc: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, message = FALSE)

library(tidyverse)
library(here)
library(vroom)
```

```{r import_ssc_to_mmu_orthologous_ids, include=FALSE}

ssc_outlier_genes_to_mmu <- here("00_analysis/data", "SOSFERT_samples_FST_analysis_genes_outliers_ssc_to_mmu.csv") %>% 
  read.csv() %>% dplyr::select(-X) %>% as_tibble() %>% 
  dplyr::rename(ensembl_gene_id_ssc = ensembl_gene_id)

```

# Overlap ssc Fst outlier genes and RED mmu genes 

ssc-outliers: regions of extreme genetic differentiation between Large-White vs Pietrain (SOSFERT samples: 10 vs 10).

mmu-REDs: regions of extreme genetic differentiation (selected mouse lines vs control line). Regions in 99th percentiles plus adjacent 95th percentile regions.

```{r overlaps_with_RED_DU_genes, include=TRUE}

lapply(c("DUK","DUC","DU6","DU6P","DUhLB"), function(du_line){
  
  # import DU RDD genes
  res <- here("../../../projekte_I2-SOS-FERT/GitHub/WGS_analysis_mmu/00_dashboard/data/RED_genes", paste0("RED_genes_",du_line,".csv")) %>% 
    read.csv() %>% dplyr::select(-X) %>% as_tibble() %>% 
    dplyr::rename(ensembl_gene_id_mmu = ensembl_gene_id) %>% 
    # inner join with ssc outliers fst
    inner_join(ssc_outlier_genes_to_mmu, 
               by = c("ensembl_gene_id_mmu" = "mmusculus_homolog_ensembl_gene"))
  
  if(nrow(res) == 0){
    paste0("There are no overlaps for ", du_line)
  }else{
    res %>% 
      dplyr::select(-seqnames, -start, -end, -width, -strand, -GOBP) %>% 
      knitr::kable()
  }
  
})

```

The only overlaps occured for DUC and DU6, same gene for each ("SRSF3 - serine and arginine rich splicing factor 3 "). 

The gene is involved in "mRNA export from nucleus", "regulation of mRNA splicing, via spliceosome" and "cellular response to leukemia inhibitory factor".

