---
title: "Genome analysis dam vs sire lines"
output: 
  html_document:
    toc: 5
---

```{r libraries, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, message = FALSE)

library(vroom)
library(tidyverse)
library(GenomicRanges)
library(here)
library(knitr)

```

```{r sample_info, include=FALSE}
sample_info <- vroom(here("sample_info/internal_external_sample_info.tsv"))

```


# Intro

Runs of homozygosity were computed for each sample of the SOSFERT sample set, using bcftools.

These samples were sequenced at high coverage of around 50x, so the VCF file had ~15M SNPs. These were used to compute ROH.

For more details check the corresponding script: in ~/12_LD_ROH/scripts/12_bcftools_ROH_SOSFERT_vcf.sh

```{r ROH_import_regions, include=FALSE}

roh_dat <- vroom(here("12_LD_ROH/output/cohort_biallelicSNPs_HardFiltered_WithMissingness_SOSFERTsamples_GTfiltered.roh.table"), 
                 col_types = c(`[3]Chromosome`="c")) %>% 
  dplyr::rename(RG = `# RG`, sample = `[2]Sample`, chr = `[3]Chromosome`, start = `[4]Start`, end = `[5]End`, length_bp = `[6]Length (bp)`,
                n_markers = `[7]Number of markers`, qual_avg_fwd_bwd_phred = `[8]Quality (average fwd-bwd phred score)`) %>% 
  filter(RG != "# RG") %>% 
  # exclude x-chrom
  filter(chr != "X") %>% 
  inner_join(
    
    # include an ordered-naming for later when plotting all samples together (i.e. boxplot)
    sample_info %>% 
      filter(project == "SOSFERT") %>% 
      group_by(breed) %>% 
      mutate(breed = ifelse(breed == "Pietrain", "PI", breed),
             breed = ifelse(breed == "Large White", "LW", breed)) %>% 
      arrange(breed, sample_id), 

    by = c("sample" = "sample_id")
    )


dim(roh_dat) # 100140     10
```

```{r classify_rohs}
# Based on: "ROHs were separated into three size classes: 1) small (<100 Kbp), 2) medium (0.1 to 5 Mbp) and 3) large (>5 Mbp). "
# PLoS Genet. 2012;8(11):e1003100. doi: 10.1371/journal.pgen.1003100. Epub 2012 Nov 29. Regions of homozygosity in the porcine genome: consequence of demography and the recombination landscape
roh_dat_incl_class <- roh_dat %>% 
  mutate(roh_class = ifelse(length_bp < 1e5, "short", NA),
         roh_class = ifelse(length_bp >= 1e5 & length_bp <= 5e6, "medium", roh_class),
         roh_class = ifelse(length_bp > 5e6, "large", roh_class),
         roh_class = factor(roh_class, levels = c("short", "medium", "large")))

```

# ROH number and size summary
```{r roh_number_size_summary}

roh_dat_incl_class %>%
  group_by(breed, roh_class) %>% 
  summarise(n = n(), min_len = min(length_bp), max_len = max(length_bp)) %>% 
  arrange(roh_class, breed) %>% 
  kable(format.args = list(big.mark = ","))
  

```


# ROH length distribution by ROH-class
Based on: "ROHs were separated into three size classes: 1) small (<100 Kbp), 2) medium (0.1 to 5 Mbp) and 3) large (>5 Mbp). "
PLoS Genet. 2012;8(11):e1003100. doi: 10.1371/journal.pgen.1003100. Epub 2012 Nov 29. Regions of homozygosity in the porcine genome: consequence of demography and the recombination landscape

```{r roh_length_boxplots, fig.width=10}

roh_dat_incl_class %>% 
  ggplot(aes(x = length_bp, y = sample)) +
    geom_boxplot() +
    facet_grid(breed ~ roh_class, scales = "free") +
    theme_bw() +
    scale_x_continuous(breaks = scales::pretty_breaks(5)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5))

```

# Number of roh per class in each sample

```{r summarise_roh_counts}

roh_dat_incl_class %>% 
  group_by(sample,breed,roh_class) %>% 
  summarise(n=n()) %>% 
  reshape2::dcast(breed+sample ~ roh_class, value.var = "n") %>% 
  kable()
  
```

# Agreggate lengths (bp) of roh per class in each sample
```{r summarise_roh_aggreg_len}

roh_dat_incl_class %>% 
  group_by(sample,breed,roh_class) %>% 
  summarise(total_length=sum(length_bp)) %>% 
  reshape2::dcast(breed+sample ~ roh_class, value.var = "total_length") %>% 
  kable(format.args = list(big.mark = ","))

```

```{r vis_aggreagate_lengths}

roh_dat_incl_class %>% 
  group_by(breed, sample, roh_class) %>% 
  summarise(total_bp = sum(length_bp)) %>% 
  ungroup() %>% 
  group_by(breed, roh_class) %>% 
  summarise(mean = mean(total_bp/1e6), sd = sd(total_bp/1e6)) %>% 
  # add length of roh for reference
  mutate(
    roh_class = as.character(roh_class),
    roh_class = ifelse(roh_class == "large", "large (>5Mb)", roh_class),
    roh_class = ifelse(roh_class == "medium", "medium (100Kb-5Mb)", roh_class),
    roh_class = ifelse(roh_class == "short", "short (<100Kb)", roh_class),
    roh_class = factor(roh_class, levels = c("total","short (<100Kb)", "medium (100Kb-5Mb)", "large (>5Mb)"))
  ) %>% 
  ggplot(aes(x = breed, y = mean, fill = roh_class)) +
    geom_bar(stat = "identity", position = "dodge") +
    #geom_bar(stat = "identity", position = "stack") +
    geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd), position =  position_dodge2(width = 0.5, padding = 0.5)) +
    theme_bw(base_size = 12) +
    xlab(NULL) +
    ylab("Mbp (mean +/- sd)") +
    ggtitle("Genome in ROH") +
    scale_fill_brewer(palette="Dark2") +
    scale_y_continuous(breaks = scales::pretty_breaks(10)) +
    theme(legend.title = element_blank(), plot.title = element_text(hjust = 0.5))


```


# Total fraction of genome in ROH (FROH)

The length of the genome covered by SNPs (denominator) is the result of:
"summing the distance between the first and the last available consecutive SNP on each chromosomal arm for each of the 22 autosomes" https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4573243/

Unfortunately, for pig centromeres are not mapped. The next best thing is to ignore per-arm territory and use the territory covered by all SNPs in the chromosome. 

```{r snp_territory}

genome_length <- here("10_FinalVCF/output/cohort_biallelicSNPs_HardFiltered_WithMissingness_SOSFERTsamples_GTfiltered.frq") %>% 
  vroom(col_select = c("CHROM","POS")) %>% 
  filter(CHROM %in% 1:18) %>% 
  group_by(CHROM) %>%
  summarise(territory_bp = (max(POS) - min(POS))+1 ) %>% 
  ungroup() %>% 
  summarise(territory_bp = sum(territory_bp)) %>% 
  as.numeric()
  
```

```{r chr_len}

len_assembly <- here("ref_ensemble95/Sus_scrofa.Sscrofa11.1.dna.toplevel.fa.fai") %>% vroom(col_names = FALSE) %>% summarise(sum(X2))

len_assembly_autosomes <- here("ref_ensemble95/Sus_scrofa.Sscrofa11.1.dna.toplevel.fa.fai") %>% 
  vroom(col_names = FALSE) %>% 
  filter(X1 %in% c(1:18)) %>% 
  summarise(sum(X2))

```

The genome length covered by SNPs (autosomes only) is `r genome_length`, whereas the length of the whole pig assembly is `r len_assembly`, only autosomes: `r len_assembly_autosomes`


```{r bp_in_roh_and_FROH_by_sample}

roh_dat_incl_class %>% 
  group_by(sample,breed) %>% 
  summarise(total_length_bp=sum(length_bp)) %>% 
  mutate(FROH = total_length_bp/genome_length) %>% 
  reshape2::melt(id.vars = c("sample","breed"), meassure.vars = c("total_length_bp", "FROH")) %>% 
  ggplot(aes(x = sample, y = value)) +
    geom_bar(stat = "identity") +
    facet_grid(variable ~ breed, scales = "free") +
    theme_bw(base_size = 12) +
    theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5)) +
    scale_y_continuous(breaks = scales::pretty_breaks(10)) +
    ylab(NULL) + 
    xlab(NULL)

```


```{r per_sample_avg_genome_prop_as_roh} 

# get genomic fraction
roh_dat_incl_class_genome_fraction <- roh_dat_incl_class %>% 
  mutate(genome_fraction = length_bp/genome_length)

# get total fraction per sample
total <- roh_dat_incl_class_genome_fraction %>% 
  group_by(breed, sample) %>% 
  summarise(total = sum(genome_fraction)) %>% 
  mutate(roh_class = "total") %>% 
  dplyr::select(breed, sample, roh_class, total)

# get total fraction by class per sample
total_by_class <- roh_dat_incl_class_genome_fraction %>% 
  group_by(breed, sample, roh_class) %>% 
  summarise(total = sum(genome_fraction))

# combine data
fractions <- bind_rows(total, total_by_class) %>% 
  #summarise by breed and class
  group_by(breed, roh_class) %>% 
  summarise(n(), mean = mean(total), sd = sd(total)) %>% 
  # add length of roh for reference
  mutate(
    roh_class = ifelse(roh_class == "large", "large (>5Mb)", roh_class),
    roh_class = ifelse(roh_class == "medium", "medium (100Kb-5Mb)", roh_class),
    roh_class = ifelse(roh_class == "short", "short (<100Kb)", roh_class)
  ) %>% 
  # arrange levels for plot
  mutate(roh_class = factor(roh_class, levels = c("total","short (<100Kb)", "medium (100Kb-5Mb)", "large (>5Mb)")))


errors_lower <- fractions$mean - fractions$sd
errors_upper <- fractions$mean + fractions$sd

ggplot(fractions, aes(x = breed, y = mean, fill = roh_class)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = errors_lower, ymax = errors_upper), position =  position_dodge2(width = 0.5, padding = 0.5)) +
  theme_bw(base_size = 12) +
  xlab(NULL) +
  ylab("FROH (mean +/- sd)") +
  ggtitle("Mean FROH per breed per ROH class") +
  scale_fill_brewer(palette="Dark2") +
  scale_y_continuous(breaks = seq(0,1,0.05), expand = c(0.01,0.01)) +
  theme(legend.title = element_blank(), plot.title = element_text(hjust = 0.5))

```

# Private and shared ROHs

- Within breed common ROHs (breed-consensus) were detected using bedtools (see script: "./12_LD_ROH/scripts/14_bedtools_multiIntersectBed.sh").

- Based on biostars post: https://www.biostars.org/p/15303/

- ROHs are transformed to BED files for bedtools, afterwards intervals were reconverted into 1-start-fully-closed intervals.

- For more information see https://genome-blog.soe.ucsc.edu/blog/2016/12/12/the-ucsc-genome-browser-coordinate-counting-systems/

```{r transform_to_bed, eval=FALSE}

roh_dat_incl_class$sample %>% 
  unique() %>% 
  lapply(function(s){
    
    d <- roh_dat_incl_class %>% 
      filter(sample == s) %>% 
      # to transform to bed, substract 1bp to start only. end stays the same.
      mutate(start = start - 1) 
    
    breed <- unique(d$breed) %>% unique()
    
    fl_nm <- paste0(breed,"_",s,".bed")
    
    d %>% 
      dplyr::select(chr, start, end) %>% 
      arrange(chr, start) %>% 
      write.table(file = here("00_analysis/data/bed_files_ROH",fl_nm), sep = "\t", quote = FALSE, col.names = FALSE, row.names = FALSE)
  })

```

```{r import_bed_and_count_intersections}
roh_multint_dat <- here("12_LD_ROH/output/SOSFERT_ROH_intersections.bed") %>% 
  vroom(col_names = FALSE) %>% 
  dplyr::rename(seqnames=X1, start=X2, end=X3) %>% 
  # do not forget to convert back to 1-start-fully-closed format
  # just add 1bp to start
  mutate(start = start + 1) 
```


```{r process_bed_and_count_intersections, eval = FALSE}

breed_cts <- roh_multint_dat$X5 %>% 
  str_split(",") %>% 
  lapply(function(x) str_sub(string = x, start = 1,end = 2)) %>% 
  lapply(function(x){
    data.frame(LW = sum(x == "LW"), PI = sum(x == "PI"))
  }) %>% 
  bind_rows()
  
roh_multint_dat2 <- bind_cols(
  
  roh_multint_dat %>% dplyr::select(seqnames, start, end),
  
  breed_cts
) 

write.csv(x = roh_multint_dat2, file = here("00_analysis/data/roh_intersections.csv"))

rm(roh_multint_dat, breed_cts, roh_multint_dat2)

```


```{r import_intersections}
roh_multint_dat2 <- here("00_analysis/data/roh_intersections.csv") %>% vroom() %>% dplyr::select(-"...1")
```

## Breed-private consensus ROH

```{r vis_bp_private_consensus_roh}

total_bp_private_LW <- roh_multint_dat2 %>% 
  filter(LW > 0, PI == 0) %>% 
  mutate(length = (end-start)+1) %>% 
  group_by(LW) %>% 
  summarise(bp_roh = sum(length)) %>% 
  dplyr::rename(n_samples = LW)
  
total_bp_private_PI <- roh_multint_dat2 %>% 
  filter(LW == 0, PI > 0) %>% 
  mutate(length = (end-start)+1) %>% 
  group_by(PI) %>% 
  summarise(bp_roh = sum(length)) %>% 
  dplyr::rename(n_samples = PI)

dat_vis_private <- list(bp_private_LW = total_bp_private_LW,
     bp_private_PI= total_bp_private_PI) %>% 
  bind_rows(.id="data_set") %>% 
  mutate(intersection_class = ifelse(n_samples <= 5, "1-5", "6-10")) %>% 
  reshape2::melt(id.vars = c("data_set", "n_samples", "bp_roh")) %>% 
  arrange(data_set, desc(n_samples)) %>% 
  group_by(data_set) %>% 
  mutate(cumsum = cumsum(bp_roh)) 

```

```{r tabulate_private_roh_bps_in_at_least_n_samples}

dat_vis_private %>% 
  reshape2::dcast(n_samples ~ data_set, value.var = "cumsum") %>% 
  dplyr::rename(min_no_samples = n_samples) %>% 
  kable(format.args = list(big.mark = ","), caption = "Total breed-private bp in ROH in at least n samples")

```

## Shared ROHs between LW and PI

```{r vis_shared_roh_bps_in_at_least_n_samples}

dat_vis_shared <- lapply(1:10, function(i){
  
  roh_multint_dat2 %>% 
    filter(LW == i & PI == i) %>% 
    mutate(n_samples = LW + PI,
           samples_per_breed = i) 
  
  }) %>% 
  bind_rows() %>% 
  group_by(samples_per_breed) %>% 
  mutate(length = (end-start)+1) %>% 
  summarise(total_bp = sum(length)) %>% 
  arrange(desc(samples_per_breed)) %>% 
  mutate(cumsum = cumsum(total_bp)) %>%
  mutate(intersection_class = ifelse(samples_per_breed <= 5, "1-5", "6-10"))
  
```

```{r tabulate_shared_roh_bps_in_at_least_n_samples}

dat_vis_shared %>% 
  arrange(samples_per_breed) %>% 
  dplyr::select(samples_per_breed, cumsum) %>% 
  dplyr::rename(min_no_samples_per_breed = samples_per_breed, total_bp = cumsum) %>% 
  kable(format.args = list(big.mark = ","), caption = "Total shared bp in ROH in at least n samples")

```

## Summary 

```{r vis_private_shared_roh, fig.height=9}

# private
p1 <- dat_vis_private %>% 
  filter(value == "1-5") %>% 
  ggplot(aes(x = n_samples, y = cumsum/1e6)) +
    geom_bar(stat = "identity") +
    facet_grid(data_set~value, scales = "free") +
    scale_x_continuous(breaks = scales::pretty_breaks(5)) +
    scale_y_continuous(breaks = scales::pretty_breaks(10)) +
    theme_bw(base_size = 10) +
    ylab("Mbp") +
    xlab("min. no. samples")


p2 <- dat_vis_private %>% 
  filter(value == "6-10") %>% 
  ggplot(aes(x = n_samples, y = cumsum/1e6)) +
    geom_bar(stat = "identity") +
    facet_grid(data_set~value, scales = "free") +
    scale_x_continuous(breaks = scales::pretty_breaks(5)) +
    scale_y_continuous(breaks = scales::pretty_breaks(10)) +
    theme_bw(base_size = 10) +
    ylab("Mbp") +
    xlab("min. no. samples")

# shared
p3 <- dat_vis_shared %>% 
  filter(intersection_class == "1-5") %>% 
  ggplot(aes(x = samples_per_breed, y = cumsum/1e6)) +
    geom_bar(stat = "identity") +
    facet_wrap(~intersection_class, scales = "free") +
    scale_x_continuous(breaks = scales::pretty_breaks(5)) +
    scale_y_continuous(breaks = scales::pretty_breaks(10)) +
    theme_bw(base_size = 10) +
    ylab("Mbp") +
    xlab("min. no. samples per breed")

p4 <- dat_vis_shared %>% 
  filter(intersection_class == "6-10") %>% 
  ggplot(aes(x = samples_per_breed, y = cumsum/1e6)) +
    geom_bar(stat = "identity") +
    facet_wrap(~intersection_class, scales = "free") +
    scale_x_continuous(breaks = scales::pretty_breaks(5)) +
    scale_y_continuous(breaks = scales::pretty_breaks(10)) +
    theme_bw(base_size = 10) +
    ylab("Mbp") +
    xlab("min. no. samples per breed")

p5 <- ggpubr::ggarrange(p1,p2,ncol = 2) %>% 
  ggpubr::annotate_figure(top = "Genome in private ROH")

p6 <- ggpubr::ggarrange(p3,p4,ncol = 2) %>% 
  ggpubr::annotate_figure(top = "Genome in shared ROH")

ggpubr::ggarrange(p5,p6, nrow = 2, heights = c(1.4,1))
```

# Check sample contribution at each level of consensus

Min. no. samples is expected to vary along the genome. Check the overall contribution of each sample at any minimum sample required.

## Contribution of samples to breed-specific ROHs
```{r bp_sample_contribution_private_LW}

cols <- c("seqnames", "start", "end", "LW_H29777", "LW_H29778", "LW_H29779", "LW_H29780", "LW_H29781", "LW_H29782", "LW_H29783", "LW_H29784", "LW_H29785", "LW_H29786", "PI_H29767", "PI_H29768", "PI_H29769", "PI_H29770", "PI_H29771", "PI_H29772", "PI_H29773", "PI_H29774", "PI_H29775", "PI_H29776")

roh_multint_dat %>% 
  dplyr::select(seqnames, start, end, X6:X25) %>% 
  magrittr::set_colnames(cols) %>% 
  inner_join(
    roh_multint_dat2 %>% filter(LW > 0, PI == 0),
    by = c("seqnames","start","end")
  ) %>% 
  reshape2::melt(id.vars = c("seqnames","start","end","LW","PI"), variable.name = "sample_id") %>% 
  as_tibble() %>% 
  filter(value > 0) %>% 
  group_by(LW, sample_id) %>% 
  mutate(length_bp = (end-start) + 1) %>% 
  summarise(total_Mbp = sum(length_bp)/1e6) %>% 
  mutate(sample_id = str_remove(sample_id, "LW_")) %>% 
  ggplot(aes(x = sample_id, y = total_Mbp)) +
    geom_bar(stat = "identity") +
    facet_wrap(~LW, scales = "free_y") +
    theme_bw(base_size = 12) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    ggtitle("bp_sample_contribution_private_LW")

```

```{r bp_sample_contribution_private_PI}
roh_multint_dat %>% 
  dplyr::select(seqnames, start, end, X6:X25) %>% 
  magrittr::set_colnames(cols) %>% 
  inner_join(
    roh_multint_dat2 %>% filter(LW == 0, PI > 0),
    by = c("seqnames","start","end")
  ) %>% 
  reshape2::melt(id.vars = c("seqnames","start","end","LW","PI"), variable.name = "sample_id") %>% 
  as_tibble() %>% 
  filter(value > 0) %>% 
  group_by(PI, sample_id) %>% 
  mutate(length_bp = (end-start) + 1) %>% 
  summarise(total_Mbp = sum(length_bp)/1e6) %>% 
  mutate(sample_id = str_remove(sample_id, "PI_")) %>% 
  ggplot(aes(x = sample_id, y = total_Mbp)) +
    geom_bar(stat = "identity") +
    facet_wrap(~PI, scales = "free_y") +
    theme_bw(base_size = 12) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    ggtitle("bp_sample_contribution_private_PI")
```

```{r bp_sample_contribution_shared, fig.width=15}

roh_multint_dat %>% 
  dplyr::select(seqnames, start, end, X6:X25) %>% 
  magrittr::set_colnames(cols) %>% 
  inner_join(
    
    lapply(1:10, function(i){
      roh_multint_dat2 %>% 
        filter(LW == i & PI == i) %>% 
        mutate(n_samples = LW + PI,
               samples_per_breed = i) 
      }) %>% 
      bind_rows() %>% 
      dplyr::select(-n_samples, -LW, -PI),
    
    by = c("seqnames","start","end")
    
  ) %>% 
  reshape2::melt(id.vars = c("seqnames","start","end","samples_per_breed"), variable.name = "sample_id") %>% 
  as_tibble() %>% 
  filter(value > 0) %>% 
  group_by(samples_per_breed, sample_id) %>% 
  mutate(length_bp = (end-start) + 1) %>% 
  summarise(total_Mbp = sum(length_bp)/1e6) %>% 
  mutate(sample_id = str_remove(sample_id, "PI_"),
         sample_id = str_remove(sample_id, "LW_")) %>% 
  ggplot(aes(x = sample_id, y = total_Mbp)) +
    geom_bar(stat = "identity") +
    facet_wrap(~samples_per_breed, scales = "free_y") +
    theme_bw(base_size = 12) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    ggtitle("bp_sample_contribution_shared")

```

